A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-15 Manuscript/OldVsns/Manuscript-7-MLC.tex
pnas vsn

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers

@@

In PGLS analysis, \(\lambda\) is a parameter that is tuned via
log-likelihood profiling for its estimation. After obtaining an
optimized \(\lambda\) value, \(\hat{\lambda}\), likelihood ratio tests
can be performed to compare observed model fits using \(\hat{\lambda}\)
to those obtained when \(\lambda=0\) or \(\lambda=1\) to infer whether
phylogenetic signal differs from no signal or a Brownian motion (BM)
model of evolutionary divergence, respectively (Freckleton et al. 2002;
Cooper et al. 2010; Bose et al. 2019). Confidence limits on
\(\hat{\lambda}\) can be also estimated based on percentiles of a
\(\chi^2\) distribution and where these values are mapped onto a
log-likelihood profile (\emph{sensu} Freckleton et al. 2002). With
respect to confidence limits, one can ascertain whether \(\lambda=0\) or
\(\lambda=1\) is contained within an interval of \(\hat{\lambda}\), as
an analog to likelihood ratio tests (see, e.g., Vandelook et al. 2019).
It is, therefore, tempting to regard \(\hat{\lambda}\) as a descriptive
statistic that measures the relative strength of phylogenetic signal,
for inferring the extent to which shared evolutionary history has
influenced trait covariation among taxa, is common. The appeal of
\(\hat{\lambda}\) as a descriptive statistic is that it serves as the
basis for interpreting ``weak'' versus ``strong'' phylogenetic signal;
i.e., small versus large values of \(\hat{\lambda}\), respectively, in a
comparative sense (e.g., De Meester et al. 2019; Pintanel et al. 2019;
Su et al. 2019). For example, qualitative comparisons of
\(\hat{\lambda}\) for multiple traits on the same phylogenetic tree
provide a basis for inferring whether the strength of phylogenetic
signal is greater in one trait as compared to another (e.g., Liu et al.
2019; Bai et al. 2019). Indeed, statements regarding the strength of
phylogenetic signal based on \(\hat{\lambda}\) are rather common in the
evolutionary literature. For instance, of the 204 papers published in
2019 that estimated and reported Pagel's \(\lambda\) (found from a
literature survey we conducted in Google.scholar), 40\% explicitly
interpreted the strength of phylogenetic signal for at least one
phenotypic trait. Further, because nearly half of the 1,572
\(\hat{\lambda}\) values reported were near 0 or 1 (Figure 1) where the
biological interpretation of \(\lambda\) is known (\(\lambda=0\)
represents no phylogenetic signal, while \(\lambda=1\) is phylogenetic
signal as expected under BM), this percentage is even higher.
\hfill\break

@@

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\hat{\lambda}\), as \(\lambda\) is a parameter
on a bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. However, equating values of \(\hat{\lambda}\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\hat{\lambda}\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. (In order to be precise, the
log-likelihood profile would have to be peaked rather than flat.)
Therefore, understanding the precision in estimating \(\hat{\lambda}\)
is paramount. One study (Boettiger et al. 2012) found that estimates of
Pagel's \(\lambda\) displayed less variation (i.e., greater precision)
when data were simulated on a large phylogeny (\(N=281\)) as compared to
a small one (\(N=13\)). From this observation it was concluded that
insufficient data (i.e., low number of species) was the underlying cause
of the increased variation across parameter estimates (Boettiger et al.
2012). Indeed, such a pattern is common with statistical estimators, as
summary statistics and parameters are often more precise at greater
sample sizes (Cohen 1988). However, this conclusion could imply that the
shape of log-likelihood profiles are assumed consistent with species
number. Such an assumption and the importance of tree shape have to date
not been verified. Thus, despite widespread use of Pagel's (1999)
\(\lambda\) in macroevolutionary studies, at present, we lack a general
understanding of the precision with which \(\hat{\lambda}\) can estimate
levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values, \(\hat{\lambda}\), of this parameter correspond to the actual
strength of the underlying input signal in the data. For this to be the
case, \(\hat{\lambda}\) must be a statistical effect size. Effect sizes
are a measure of the magnitude of a statistical effect in data,
represented on a common scale (Glass 1976; Cohen 1988). Effect sizes
have widespread use in many areas of the quantiative sciences, as they
represent measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunately, not all model parameters and descriptive statistics are
effect sizes, directly, and thus many summary measures must first be
converted to statistics with standardized units (i.e., conversion to an
effect size) for meaningful comparison (see Rosenthal 1994). As a
consequence, it follows that only if \(\hat{\lambda}\) is a statistical
effect size can comparisons of estimates across datasets be
interpretable. However, the calculation and statistical behavior of
\(\hat{\lambda}\) as an effect size has not yet been explored.
\hfill\break

As a prortional random variable bounded by 0 and 1, we might expect that
\(\hat{\lambda}\) follows A Bernoulli distribution; i.e., branch lengths
in a tree are scaled proportionally to the probability that data arise
from a BM process. As such, we would also expect in simulation
experiment that given a known \(\lambda\) value used to generate random
data on a tree: the mean of an empirical sampling distribution of
\(\hat{\lambda}\) would approximately equal \(\lambda\); the standard
error of \(\hat{\lambda}\) would be largest at intermediate values of
\(\lambda\), and predictable over the range of \(\lambda\) with respect
to treesize; the distribution of \(\hat{\lambda}\) would be symmetric at
intermediate values of \(\lambda\) and more skewed toward values of 0 or
1; and that the distribution of \(\hat{\lambda}\) will be more
platykurtic at intermediate values of \(\lambda\), becoming more
leptokurtic toward 0 and 1. These properties seemed to be
superficiallyreasonable from the simulations perfomed by Münkemüller et
al.(2012, see Fig. 2). Their study simulated strength of phylogenetic
signal as a varied weighted-average of data simulated on trees with
\(\lambda=0\) and \(\lambda=1\), rather than from prescribed values of
\(\lambda\). Nonetheless, their results revealed that as a random
variable, \(\hat{\lambda}\) resembled a Bernoulli random variable based
superficially on statistical moments (mean, variance, skewness, and
kurtosis; see their Fig. 2) with respect to strength of phylogenetic
signal, for a given tree size. However, for comparisons among tree sizes
for a given strength of phylogenetic signal, median values of
\(\hat{\lambda}\) were slightly, positively associated with the number
of species in a tree, especially at weaker strength of signals,
suggesting \(\hat{\lambda}\) might not be an accurate statistic across a
range of tree sizes. In order for \(\hat{\lambda}\) to be considered a
reliable statistic as a measure of phylogenetic signal, as a minimum, it
should be accurate and precise, properties that could be measured from
simulation experiments for known values of \(\lambda\).

@@

\hypertarget{the-precision-of-hatlambda-is-variable}{%
\subsection{\texorpdfstring{\emph{The Precision of \(\hat{\lambda}\) is
Variable}}{The Precision of \textbackslash{}hat\{\textbackslash{}lambda\} is Variable}}\label{the-precision-of-hatlambda-is-variable}}

We conducted a series of computer simulations to evaluate the precision
of \(\hat{\lambda}\) based on varied input of Pagel's \(\lambda\) to
generate data. Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). For each simulation run, we rescaled the
simulated phylogenies by multiplying the internal branches by
\(\lambda_{in}\), using 21 intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a BM model of evolution to obtain
datasets with differing levels of phylogenetic signal, that ranged from
no phylogenetic signal (when \(\lambda_{in} =0\)), to phylogenetic
signal reflecting a BM model of evolution (when \(\lambda_{in} =1\)).
For each dataset we then estimated phylogenetic signal
(\(\hat{\lambda}\)), and calculated its variance
(\(\sigma^2_{\hat{\lambda}}\)) at each input level of phylogenetic
signal and level of species richness as an estimate of precision. Random
data samples were generated from a normal distribution, i.e.,
\(\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})\). We verified that the
level of \(\sigma^2\) had no effect on \(\hat{\lambda}\) or
\(\sigma^2_{\hat{\lambda}}\), but the results we present used
\(\sigma^2=1\). \hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al. (2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
published in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these mid-range empirical estimates were explicitly interpreted
in terms of the strength of phylogenetic signal that they represented
(i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

@@

This study is not the first to compare \(\lambda\) and \(\kappa\) for
their ability as statistics to measure phylogenetic signal. Our results
for \(\lambda\) and \(\kappa\) values are consistent with those found in
the simulations performed by Münkemüller et al. (2012), but that study
investigated type I error rates and statistical power, finding that
\(\lambda\) performed better in both regards, irrespective of species
number in trees. Although not the central focus of their study, the same
tendency for variable \(\lambda\) and consistent \(\kappa\) at
intermediate phylogenetic signal strengths was observed (see Fig. 2,
Münkemüller et al. 2012). Recent work by Molina-Venegas and Rodríguez
(2017) found that \(\kappa\) but not \(\lambda\) tended to inflate the
estimate of phylogenetic signal, leading to moderate type I and type II
biases, if polytomic chronograms were used. Their work more thoroughly
addressed previous observations of inflated \(\kappa\) for incompletely
resolved phylogenetic trees (Davies et al. 2012; Münkemüller et al.
2012). An interesting question is whether an inflated \(\kappa\) value
leads to an inflated \(Z_\kappa\) or does a tendency of a particular
tree to inflate estimates of \(\kappa\) also inflate the values in
random permutations of a test, in which case \(Z_\kappa\) is robust to
polytomies? We repeated the analyses in Figure 4, adjusting trees to
have 50\% collapsed nodes, per the technique of Molina-Venegas and
Rodríguez (2017), and found results were consistent (see Supporting
Information). This confirms that any tendency of incompletely resolved
trees to inflate \(\kappa\) as a descriptive statistic does not inflate
\(Z_\kappa\) as an effect size. Furthermore, because comparison of
effect sizes in a test is a comparison of locations of observed values
in their sampling distributions, which would shift concomitantly because
of this tendency, the \(Z_{12}\) test statistic in equation 4 appears to
be robust in spite of unresolved trees. \hfill\break

@@

\leavevmode\hypertarget{ref-AdamsGeomorph}{}%
Adams, D. C., M. L. Collyer, and A. Kaliontzopoulou. 2020. Geomorph:
Software for geometric morphometric analyses. R package version 3.2.1.

@@

\leavevmode\hypertarget{ref-Orme2013}{}%
Orme, D., R. Freckleton, G. Thomas, T. Petzoldt, S. Fritz, and N. Isaac.
2013. CAPER: Comparative analyses of phylogenetics and evolution in r.
Methods in Ecology and Evolution 3:145--151.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers

@@

In PGLS analysis, \(\lambda\) is a parameter that is tuned via
log-likelihood profiling for its estimation. After obtaining an
optimized \(\lambda\) value, \(\hat{\lambda}\), likelihood ratio tests
can be performed to compare observed model fits using \(\hat{\lambda}\)
to those obtained when \(\lambda=0\) or \(\lambda=1\) to infer whether
phylogenetic signal differs from no signal or a Brownian motion (BM)
model of evolutionary divergence, respectively (Freckleton et al. 2002;
Cooper et al. 2010; Bose et al. 2019). Confidence limits on
\(\hat{\lambda}\) can also be estimated based on percentiles of a
\(\chi^2\) distribution and where these values are mapped onto a
log-likelihood profile (\emph{sensu} Freckleton et al. 2002). With
respect to confidence limits, one can ascertain whether \(\lambda=0\) or
\(\lambda=1\) is contained within an interval of \(\hat{\lambda}\), as
an analog to likelihood ratio tests (see, e.g., Vandelook et al. 2019).
It is, therefore, tempting to regard \(\hat{\lambda}\) as a descriptive
statistic that measures the relative strength of phylogenetic signal,
for inferring the extent to which shared evolutionary history has
influenced trait covariation among taxa. The appeal of \(\hat{\lambda}\)
as a descriptive statistic is that it serves as the basis for
interpreting ``weak'' versus ``strong'' phylogenetic signal; i.e., small
versus large values of \(\hat{\lambda}\), respectively, in a comparative
sense (e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al.
2019). For example, qualitative comparisons of \(\hat{\lambda}\) for
multiple traits on the same phylogenetic tree provide a basis for
inferring whether the strength of phylogenetic signal is greater in one
trait as compared to another (e.g., Liu et al. 2019; Bai et al. 2019).
Indeed, statements regarding the strength of phylogenetic signal based
on \(\hat{\lambda}\) are rather common in the evolutionary literature.
For instance, of the 204 papers published in 2019 that estimated and
reported Pagel's \(\lambda\) (found from a literature survey we
conducted in Google.scholar), 40\% explicitly interpreted the strength
of phylogenetic signal for at least one phenotypic trait. Further,
because nearly half of the 1,572 \(\hat{\lambda}\) values reported were
near 0 or 1 (Figure 1) where the biological interpretation of
\(\lambda\) is known (\(\lambda=0\) represents no phylogenetic signal,
while \(\lambda=1\) is phylogenetic signal as expected under BM), this
percentage is even higher. \hfill\break

@@

\hypertarget{as-a-statistic-that-measures-phylogenetic-signal-how-should-hatlambda-behave}{%
\section{\texorpdfstring{As a statistic that measures phylogenetic
signal, how should \(\hat{\lambda}\)
behave?}{As a statistic that measures phylogenetic signal, how should \textbackslash hat\{\textbackslash lambda\} behave?}}\label{as-a-statistic-that-measures-phylogenetic-signal-how-should-hatlambda-behave}}

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\hat{\lambda}\), as \(\lambda\) is a parameter
on a bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. However, equating values of \(\hat{\lambda}\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\hat{\lambda}\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. (In order to be precise, the
log-likelihood profile would have to be peaked rather than flat.)
Therefore, understanding the precision in estimating \(\hat{\lambda}\)
is paramount. As a prortional random variable bounded by 0 and 1, we
might expect that \(\hat{\lambda}\) follows A Bernoulli distribution;
i.e., branch lengths in a tree are scaled proportionally to the
probability that data arise from a BM process. As such, we would also
expect in simulation experiment that given a known \(\lambda\) value
used to generate random data on a tree: the mean of an empirical
sampling distribution of \(\hat{\lambda}\) would approximately equal
\(\lambda\); the standard error of \(\hat{\lambda}\) would be largest at
intermediate values of \(\lambda\), and predictable over the range of
\(\lambda\) with respect to treesize; the distribution of
\(\hat{\lambda}\) would be symmetric at intermediate values of
\(\lambda\) and more skewed toward values of 0 or 1; and that the
distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1. These properties seemed to be superficially reasonable from the
simulations perfomed by Münkemüller et al.(2012, see Fig. 2). Their
study simulated strength of phylogenetic signal as a varied
weighted-average of data simulated on trees with \(\lambda=0\) and
\(\lambda=1\), rather than from prescribed values of \(\lambda\).
Nonetheless, their results revealed that as a random variable,
\(\hat{\lambda}\) resembled a Bernoulli random variable based
superficially on statistical moments (mean, variance, skewness, and
kurtosis; see their Fig. 2) with respect to strength of phylogenetic
signal, for a given tree size. However, for comparisons among tree sizes
for a given strength of phylogenetic signal, median values of
\(\hat{\lambda}\) were slightly, positively associated with the number
of species in a tree, especially at weaker strength of signals,
suggesting \(\hat{\lambda}\) might not be an accurate statistic across a
range of tree sizes. \hfill\break

Another study (Boettiger et al. 2012) found that \(\hat{\lambda}\)
displayed less variation (i.e., greater precision) when data were
simulated on a large phylogeny (\(N=281\)) as compared to a small one
(\(N=13\)). From this observation it was concluded that insufficient
data (i.e., low number of species) was the underlying cause of the
increased variation across parameter estimates (Boettiger et al. 2012).
Indeed, such a pattern is common with statistical estimators, as summary
statistics and parameters are often more precise at greater sample sizes
(Cohen 1988). However, this conclusion could imply that the shape of
log-likelihood profiles are assumed consistent with species number. Such
an assumption and the importance of tree shape have to date not been
verified. Thus, despite widespread use of Pagel's (1999) \(\lambda\) in
macroevolutionary studies, at present, we lack a general understanding
of the precision with which \(\hat{\lambda}\) can estimate levels of
phylogenetic signal in phenotypic datasets. \hfill\break

Second, while \(\lambda\) varies over a bounded scale (\(0\to1\)), this
does not \emph{de-facto} imply that the estimated values,
\(\hat{\lambda}\), of this parameter correspond to the actual strength
of the underlying input signal in the data. For this to be the case,
\(\hat{\lambda}\) must be a statistical effect size. Effect sizes are a
measure of the magnitude of a statistical effect in data, represented on
a common scale (Glass 1976; Cohen 1988). Effect sizes have widespread
use in many areas of the quantiative sciences, as they represent
measures that may be readily summarized across datasets as in
meta-analyses (Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster
1995), or compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunately, not all model parameters and descriptive statistics are
effect sizes, directly, and thus many summary measures must first be
converted to statistics with standardized units (i.e., conversion to an
effect size) for meaningful comparison (see Rosenthal 1994). As a
consequence, it follows that only if \(\hat{\lambda}\) can be cobverted
to a statistical effect size can comparisons of estimates across
datasets be interpretable. However, the calculation and statistical
behavior of \(\hat{\lambda}\) as an effect size has not yet been
explored. \hfill\break

@@

\hypertarget{the-accuracy-and-precision-of-hatlambda-is-variable}{%
\subsection{\texorpdfstring{\emph{The accuracy and precision of
\(\hat{\lambda}\) is
Variable}}{The accuracy and precision of \textbackslash hat\{\textbackslash lambda\} is Variable}}\label{the-accuracy-and-precision-of-hatlambda-is-variable}}

We conducted a series of computer simulations to evaluate the precision
of \(\hat{\lambda}\) based on varied input of Pagel's \(\lambda\) to
generate data. Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). For each simulation run, we rescaled the
simulated phylogenies by multiplying the internal branches by
\(\lambda_{in}\), using 21 intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a BM model of evolution to obtain
datasets with differing levels of phylogenetic signal, that ranged from
no phylogenetic signal (when \(\lambda_{in} =0\)), to phylogenetic
signal reflecting a BM model of evolution (when \(\lambda_{in} =1\)).
For each dataset we then estimated phylogenetic signal
(\(\hat{\lambda}\)), and calculated the standard deviation of
\({\hat{\lambda}\) (\(\sigma_{\hat{\lambda}}\)) at each input level of
phylogenetic signal and level of species richness as an estimate of
precision. Random data samples were generated from a normal
distribution, i.e., \(\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})\). We
verified that the level of \(\sigma^2\) had no effect on
\(\hat{\lambda}\) or \(\sigma^2_{\hat{\lambda}}\), but the results we
present used \(\sigma^2=1\). \hfill\break

@@

\emph{Results.}

superifical Bernoulli traits innacuracy precision

We found that the precision of \(\lambda_{est}\) varied widely across
simulation conditions. Predictably, precision improved as the number of
species increased (Figure 2). This confirmed earlier findings of
Boettiger et al.~(2012), and adhered to parametric statistical theory.
However, in many cases the set of \(\lambda_{est}\) spanned nearly the
entire range of possible values (e.g., \(n=32\); \(\lambda_{in}=0.5\):
\(\lambda_{est}= 0.0\to 0.985\)), revealing that estimates of
\(\lambda\) were not a reliable indicator of input phylogenetic signal.
Importantly, the precision of \(\lambda_{est}\) was not uniform across
all levels of phylogenetic signal. The worst precision was observed at
intermediate levels of phylogenetic signal (\(\lambda_{in}\approx0.5\)),
while precision improved as input levels approached the extremes of
\(\lambda\)'s range (i.e., \(\lambda_{in}\rightarrow0\) \&
\(\lambda_{in}\rightarrow1\)). Thus, estimates of \(\lambda\) were least
reflective of the true input signal at intermediate values.
Additionally, even at large levels of species richness, we found that
the range of \(\lambda_{est}\) still encompassed a substantial portion
of possible values (e.g., \(n=512\); \(\lambda_{in}=0.5\):
\(\lambda_{est} = 0.32\to 0.68\)). Likewise, the same \(\lambda_{est}\)
could be obtained from datasets containing vastly different input levels
of phylogenetic signal (e.g., \(n=512\); \(\lambda_{est} = 0.5\);
\(\lambda_{in} = 0.25 \to 0.65\)). These findings were particularly
unsettling when considered in light of our literature survey. Over one
quarter of the \(\lambda\) estimates published in empirical studies (421
of 1,572) were between \(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1).
This range reflected the region that our simulations identified as being
the least reliable in terms of accurately characterizing levels of
phylogenetic signal, yet 30\% of these mid-range empirical estimates
were explicitly interpreted in terms of the strength of phylogenetic
signal that they represented (i.e., weak, intermediate, strong
phylogenetic signal). \hfill\break

@@

This study is not the first to compare \(\lambda\) and \(\kappa\) for
their ability as statistics to measure phylogenetic signal. Our results
for \(\lambda\) and \(\kappa\) values are consistent with those found in
the simulations performed by Münkemüller et al.~(2012), but that study
investigated type I error rates and statistical power, finding that
\(\lambda\) performed better in both regards, irrespective of species
number in trees. Although not the central focus of their study, the same
tendency for variable \(\lambda\) and consistent \(\kappa\) at
intermediate phylogenetic signal strengths was observed (see Fig. 2,
Münkemüller et al. 2012). Recent work by Molina-Venegas and Rodríguez
(2017) found that \(\kappa\) but not \(\lambda\) tended to inflate the
estimate of phylogenetic signal, leading to moderate type I and type II
biases, if polytomic chronograms were used. Their work more thoroughly
addressed previous observations of inflated \(\kappa\) for incompletely
resolved phylogenetic trees (Davies et al. 2012; Münkemüller et al.
2012). An interesting question is whether an inflated \(\kappa\) value
leads to an inflated \(Z_\kappa\) or does a tendency of a particular
tree to inflate estimates of \(\kappa\) also inflate the values in
random permutations of a test, in which case \(Z_\kappa\) is robust to
polytomies? We repeated the analyses in Figure 4, adjusting trees to
have 50\% collapsed nodes, per the technique of Molina-Venegas and
Rodríguez (2017), and found results were consistent (see Supporting
Information). This confirms that any tendency of incompletely resolved
trees to inflate \(\kappa\) as a descriptive statistic does not inflate
\(Z_\kappa\) as an effect size. Furthermore, because comparison of
effect sizes in a test is a comparison of locations of observed values
in their sampling distributions, which would shift concomitantly because
of this tendency, the \(Z_{12}\) test statistic in equation 4 appears to
be robust in spite of unresolved trees. \hfill\break

@@

\leavevmode\hypertarget{ref-AdamsGeomorph}{}%
Adams, D. C., M. L. Collyer, and A. Kaliontzopoulou. 2020. Geomorph:
Software for geometric morphometric analyses. R package version 3.3.1.

@@

\leavevmode\hypertarget{ref-Orme2013}{}%
Orme, D., R. P. Freckleton, G. H. Thomas, T. Petzoldt, S. A. Fritz, and
N. Isaac. 2013. CAPER: Comparative analyses of phylogenetics and
evolution in r. Methods in Ecology and Evolution 3:145--151.
